---
title: "Hands-on Exercise 1: Geospatial Data Wrangling and Choropleth Mapping with R"
execute:
  echo: true
  warning: false
  message: false
format:
  html:
    toc: true
    toc-depth: 3
---

## Overview

This page combines two Hands-on Exercise 1 topics:

-   [Geospatial Data Wrangling with R](https://r4gdsa.netlify.app/chap01.html)
-   [Choropleth Mapping with R](https://r4gdsa.netlify.app/chap02.html)

I followed most of the important demos from the course website.

## Part 1: Geospatial Data Wrangling with R

This part is about preparing spatial data before analysis. I need to import the data, check the CRS, convert coordinate tables into spatial points, and do simple geoprocessing.

### Load Packages

```{r}
library(sf)
library(tidyverse)
library(tmap)
library(knitr)

tmap_mode("plot")
```

### Data Used

```{r}
data_files <- tibble(
  data_type = c(
    "Geospatial",
    "Geospatial",
    "Geospatial",
    "Geospatial",
    "Aspatial",
    "Aspatial"
  ),
  file = c(
    "Master Plan 2014 Subzone Boundary (No Sea)",
    "Master Plan 2019 Subzone Boundary (No Sea)",
    "Cycling Path Network",
    "Pre-Schools Location",
    "Inside Airbnb listings",
    "Singapore resident population by subzone and age"
  ),
  path = c(
    "data/geospatial/master_plan_2014_subzone_no_sea.geojson",
    "data/geospatial/master_plan_2019_subzone_no_sea.geojson",
    "data/geospatial/cycling_path_network.geojson",
    "data/geospatial/pre_schools_location.geojson",
    "data/aspatial/listings.csv",
    "data/aspatial/respop_subzone_age_sex_2020.csv"
  )
)

kable(data_files)
```

### Import Geospatial Data

```{r}
mpsz_raw <- st_read(
  "data/geospatial/master_plan_2014_subzone_no_sea.geojson",
  quiet = TRUE
)

cyclingpath_raw <- st_read(
  "data/geospatial/cycling_path_network.geojson",
  quiet = TRUE
)

preschool_raw <- st_read(
  "data/geospatial/pre_schools_location.geojson",
  quiet = TRUE
)
```

```{r}
raw_summary <- tibble(
  layer = c("mpsz_raw", "cyclingpath_raw", "preschool_raw"),
  features = c(nrow(mpsz_raw), nrow(cyclingpath_raw), nrow(preschool_raw)),
  geometry_type = c(
    paste(unique(st_geometry_type(mpsz_raw)), collapse = ", "),
    paste(unique(st_geometry_type(cyclingpath_raw)), collapse = ", "),
    paste(unique(st_geometry_type(preschool_raw)), collapse = ", ")
  ),
  epsg = c(
    st_crs(mpsz_raw)$epsg,
    st_crs(cyclingpath_raw)$epsg,
    st_crs(preschool_raw)$epsg
  )
)

kable(raw_summary)
```

The GeoJSON files are read as `sf` objects. At this stage, the coordinates are still in longitude and latitude.

### Transform Projection

```{r}
mpsz <- st_transform(mpsz_raw, crs = 3414)
cyclingpath <- st_transform(cyclingpath_raw, crs = 3414)
preschool <- st_transform(preschool_raw, crs = 3414)
```

```{r}
projected_summary <- tibble(
  layer = c("mpsz", "cyclingpath", "preschool"),
  epsg = c(st_crs(mpsz)$epsg, st_crs(cyclingpath)$epsg, st_crs(preschool)$epsg),
  crs_name = c(st_crs(mpsz)$Name, st_crs(cyclingpath)$Name, st_crs(preschool)$Name)
)

kable(projected_summary)
```

The course uses SVY21 / Singapore TM. It is more suitable for Singapore distance and area work because the units are metres.

### Examine Simple Feature Data

```{r}
st_geometry(mpsz)[1:2]
```

```{r}
glimpse(st_drop_geometry(mpsz))
```

```{r}
mpsz %>%
  st_drop_geometry() %>%
  select(SUBZONE_N, PLN_AREA_N, REGION_N, SHAPE_1.AREA) %>%
  head(8) %>%
  kable()
```

An `sf` data frame is still a data frame. The special part is the geometry column. This makes it easy to use `dplyr` and spatial functions together.

### Quick Plot with Base R

```{r, fig.width=7, fig.height=6}
plot(st_geometry(mpsz))
```

```{r, fig.width=7, fig.height=6}
plot(mpsz["PLN_AREA_N"])
```

The first plot only shows geometry. The second plot colours the polygons by planning area. This is good for a quick check.

### Plot Multiple Layers Together

```{r, fig.width=7, fig.height=6}
plot(st_geometry(mpsz), border = "grey70")
plot(st_geometry(preschool), add = TRUE, pch = 20, col = "red", cex = 0.35)
```

The pre-school points now sit on top of the planning subzones. This tells me the projection is working.

### A Cleaner ggplot View

```{r, fig.width=8, fig.height=7}
ggplot() +
  geom_sf(data = mpsz, fill = "grey96", color = "grey65", linewidth = 0.12) +
  geom_sf(data = cyclingpath, color = "#2c7fb8", linewidth = 0.25, alpha = 0.65) +
  geom_sf(data = preschool, color = "#de2d26", size = 0.25, alpha = 0.45) +
  coord_sf() +
  labs(
    title = "Planning Subzones, Cycling Paths, and Pre-Schools",
    subtitle = "The layers align after CRS transformation",
    caption = "Data: data.gov.sg and LTA DataMall"
  ) +
  theme_minimal()
```

This plot is just a visual check. I use it to see whether points, lines, and polygons are in the correct place.

### Import Aspatial Airbnb Data

```{r}
listings <- read_csv("data/aspatial/listings.csv", show_col_types = FALSE)

list(listings)
```

The Airbnb file is not a geospatial file. But it has longitude and latitude columns, so I can convert it into a point layer.

### Convert Aspatial Data to sf

```{r}
listings_sf <- st_as_sf(
  listings,
  coords = c("longitude", "latitude"),
  crs = 4326
) %>%
  st_transform(crs = 3414)
```

```{r}
glimpse(listings_sf)
```

```{r}
tibble(
  rows_before = nrow(listings),
  rows_after = nrow(listings_sf),
  epsg = st_crs(listings_sf)$epsg
) %>%
  kable()
```

The new `geometry` column stores each Airbnb listing as a point. After this step, I can use it like other spatial data.

### Plot Airbnb Listings on Subzones

```{r, fig.width=7, fig.height=6}
plot(st_geometry(mpsz), border = "grey75")
plot(st_geometry(listings_sf), add = TRUE, pch = 20, col = "#2171b5", cex = 0.35)
```

This follows the DIY idea from the course page. It is a quick way to check if the converted points look correct.

### Geoprocessing Use Case 1: Cycling Path Buffer

```{r}
buffer_cycling <- st_buffer(
  cyclingpath,
  dist = 5,
  nQuadSegs = 30
)
```

```{r}
buffer_cycling <- buffer_cycling %>%
  mutate(AREA = st_area(geometry))

sum(buffer_cycling$AREA)
```

The buffer shows the land around the cycling paths. The distance is 5 metres, same as the demo idea in the course notes.

### Clip Buffer by One Subzone

```{r}
mpsz_selected <- mpsz %>%
  filter(SUBZONE_N == "TAMPINES WEST")

buffer_cycling_selected <- st_intersection(
  buffer_cycling,
  mpsz_selected
) %>%
  mutate(CLIPPED_AREA = st_area(geometry))
```

```{r}
tibble(
  selected_subzone = mpsz_selected$SUBZONE_N,
  number_of_buffer_parts = nrow(buffer_cycling_selected),
  clipped_area_sqm = as.numeric(sum(buffer_cycling_selected$CLIPPED_AREA)),
  clipped_area_sqkm = as.numeric(sum(buffer_cycling_selected$CLIPPED_AREA)) / 1000000
) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3))) %>%
  kable()
```

```{r, fig.width=7, fig.height=6}
plot(st_geometry(mpsz_selected), col = "grey92", border = "grey40")
plot(st_geometry(buffer_cycling_selected), add = TRUE, col = "#9ecae1", border = NA)
plot(st_geometry(cyclingpath), add = TRUE, col = "#08519c", lwd = 0.7)
```

`st_intersection()` keeps only the part of the buffer inside Tampines West. This is useful when the question is focused on one area.

### Geoprocessing Use Case 2: Count Pre-Schools by Subzone

```{r}
mpsz$`PreSch Count` <- lengths(st_intersects(mpsz, preschool))
```

```{r}
summary(mpsz$`PreSch Count`)
```

```{r}
mpsz %>%
  slice_max(`PreSch Count`, n = 5) %>%
  st_drop_geometry() %>%
  select(SUBZONE_N, PLN_AREA_N, REGION_N, `PreSch Count`) %>%
  kable()
```

The `st_intersects()` result tells which pre-school points fall inside each polygon. `lengths()` turns that list into a count.

### Derive Area and Density

```{r}
mpsz <- mpsz %>%
  mutate(
    Area = st_area(geometry),
    `PreSch Density` = `PreSch Count` / as.numeric(Area) * 1000000
  )
```

```{r}
summary(mpsz$`PreSch Density`)
```

Density is better than raw count for comparison. A small subzone and a large subzone can have the same count, but the meaning is not the same.

### Histogram of Pre-School Density

```{r, fig.width=7, fig.height=5}
hist(
  mpsz$`PreSch Density`,
  main = "Distribution of Pre-School Density",
  xlab = "Pre-schools per square kilometre",
  col = "#9ecae1",
  border = "white"
)
```

```{r, fig.width=8, fig.height=5}
mpsz %>%
  st_drop_geometry() %>%
  ggplot(aes(x = `PreSch Density`)) +
  geom_histogram(bins = 30, fill = "#3182bd", color = "white") +
  labs(
    title = "Distribution of Pre-School Density",
    x = "Pre-schools per square kilometre",
    y = "Number of subzones"
  ) +
  theme_minimal()
```

The histogram shows that many subzones have low density. A few compact residential areas have much higher density.

### Map Pre-School Density

```{r, fig.width=8, fig.height=7}
ggplot(mpsz) +
  geom_sf(aes(fill = `PreSch Density`), color = "white", linewidth = 0.08) +
  scale_fill_viridis_c(option = "C", na.value = "grey90") +
  labs(
    title = "Pre-School Density by Planning Subzone",
    subtitle = "Number of pre-schools per square kilometre",
    fill = "Centres\nper sq km"
  ) +
  theme_minimal()
```

This map is not from `tmap`, but it helps me check the derived variable before using it in later analysis.

## Part 2: Choropleth Mapping with R

This part is about turning data into thematic maps. I use dependency ratio as the mapping variable.

### Import Mapping and Population Data

```{r}
mpsz2019 <- st_read(
  "data/geospatial/master_plan_2019_subzone_no_sea.geojson",
  quiet = TRUE
) %>%
  st_transform(3414)

clean_count <- function(x) {
  parse_number(na_if(as.character(x), "-"))
}

pop_raw <- read_csv(
  "data/aspatial/respop_subzone_age_sex_2020.csv",
  show_col_types = FALSE
) %>%
  mutate(across(starts_with("Total_"), clean_count))
```

```{r}
tibble(
  layer = c("Master Plan 2019 subzones", "Census 2020 population table"),
  records = c(nrow(mpsz2019), nrow(pop_raw)),
  columns = c(ncol(mpsz2019), ncol(pop_raw))
) %>%
  kable()
```

The course website uses a population table and a subzone layer. I use the same idea here with the downloaded local files.

### Prepare Dependency Ratio

```{r}
pop2020 <- pop_raw %>%
  filter(Number != "Total", !str_detect(Number, " - Total$")) %>%
  transmute(
    SUBZONE_N = str_to_upper(Number),
    YOUNG = Total_0_4 + Total_5_9 + Total_10_14,
    ECONOMY_ACTIVE = Total_15_19 + Total_20_24 + Total_25_29 +
      Total_30_34 + Total_35_39 + Total_40_44 + Total_45_49 +
      Total_50_54 + Total_55_59 + Total_60_64,
    AGED = Total_65_69 + Total_70_74 + Total_75_79 +
      Total_80_84 + Total_85_89 + Total_90andOver,
    TOTAL = Total_Total,
    DEPENDENCY = round((YOUNG + AGED) / ECONOMY_ACTIVE, 4)
  ) %>%
  filter(!is.na(ECONOMY_ACTIVE), ECONOMY_ACTIVE > 0)
```

```{r}
pop2020 %>%
  head(10) %>%
  kable()
```

I keep `DEPENDENCY` as a ratio. This makes the map legend closer to the course demo.

### Join Population Data to Subzone Polygons

```{r}
mpsz_pop2020 <- mpsz2019 %>%
  left_join(pop2020, by = "SUBZONE_N")
```

```{r}
join_summary <- tibble(
  subzones = nrow(mpsz_pop2020),
  matched_subzones = sum(!is.na(mpsz_pop2020$DEPENDENCY)),
  unmatched_subzones = sum(is.na(mpsz_pop2020$DEPENDENCY))
)

kable(join_summary)
```

```{r}
mpsz_pop2020 %>%
  st_drop_geometry() %>%
  select(SUBZONE_N, PLN_AREA_N, REGION_N, TOTAL, YOUNG, ECONOMY_ACTIVE, AGED, DEPENDENCY) %>%
  arrange(desc(DEPENDENCY)) %>%
  slice_head(n = 10) %>%
  kable()
```

Some subzones are unmatched. This is normal because not every planning subzone is a residential area.

### Quick Choropleth with qtm

```{r, fig.width=8, fig.height=7}
qtm(mpsz_pop2020, fill = "DEPENDENCY")
```

`qtm()` is useful for a fast map. I use it when I only need a quick first look.

### Choropleth with tm_shape and tm_polygons

```{r, fig.width=8, fig.height=7}
tm_shape(mpsz_pop2020) +
  tm_polygons(fill = "DEPENDENCY")
```

This is the basic `tmap` structure. `tm_shape()` gives the spatial object. `tm_polygons()` draws the polygons.

### Use tm_fill and tm_borders

```{r, fig.width=8, fig.height=7}
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY")
```

```{r, fig.width=8, fig.height=7}
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY") +
  tm_borders()
```

```{r, fig.width=8, fig.height=7}
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY") +
  tm_borders(
    col = "grey60",
    lwd = 0.1,
    lty = "dashed"
  )
```

`tm_polygons()` is a shortcut for fill plus borders. I like the shortcut, but separating them helps me control the border style.

### Built-In Classification Methods

```{r, fig.width=8, fig.height=7}
tm_shape(mpsz_pop2020) +
  tm_polygons(
    fill = "DEPENDENCY",
    fill.scale = tm_scale_intervals(
      style = "quantile",
      n = 5
    )
  ) +
  tm_borders(fill_alpha = 0.5)
```

```{r, fig.width=8, fig.height=7}
tm_shape(mpsz_pop2020) +
  tm_polygons(
    fill = "DEPENDENCY",
    fill.scale = tm_scale_intervals(
      style = "equal",
      n = 5
    )
  ) +
  tm_borders(fill_alpha = 0.5)
```

```{r, fig.width=8, fig.height=7}
tm_shape(mpsz_pop2020) +
  tm_polygons(
    fill = "DEPENDENCY",
    fill.scale = tm_scale_intervals(
      style = "kmeans",
      n = 5
    )
  ) +
  tm_borders(fill_alpha = 0.5)
```

Different classification methods can make the same data look different. This is why map classification is a design choice, not just a technical setting.

### Custom Breaks

```{r}
summary(mpsz_pop2020$DEPENDENCY)
```

```{r, fig.width=8, fig.height=7}
dependency_breaks <- c(0, 0.30, 0.40, 0.50, 0.60, 0.80)

tm_shape(mpsz_pop2020) +
  tm_polygons(
    fill = "DEPENDENCY",
    fill.scale = tm_scale_intervals(
      breaks = dependency_breaks
    )
  ) +
  tm_borders(fill_alpha = 0.5)
```

The website uses custom breaks too. I adjusted the break values because my downloaded population table gives a lower range.

### ColourBrewer Palette

```{r, fig.width=8, fig.height=7}
tm_shape(mpsz_pop2020) +
  tm_polygons(
    fill = "DEPENDENCY",
    fill.scale = tm_scale_intervals(
      style = "quantile",
      n = 5,
      values = "brewer.greens"
    )
  ) +
  tm_borders(fill_alpha = 0.5)
```

```{r, fig.width=8, fig.height=7}
tm_shape(mpsz_pop2020) +
  tm_polygons(
    fill = "DEPENDENCY",
    fill.scale = tm_scale_intervals(
      style = "quantile",
      n = 5,
      values = "-brewer.greens"
    )
  ) +
  tm_borders(fill_alpha = 0.5)
```

The minus sign reverses the colour order. This changes which values get the darkest colour.

### Cartographic Furniture

```{r, fig.width=8, fig.height=7}
tm_shape(mpsz_pop2020) +
  tm_polygons(
    fill = "DEPENDENCY",
    fill.scale = tm_scale_intervals(
      style = "quantile",
      n = 5
    )
  ) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type = "8star", size = 2) +
  tm_scalebar() +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits(
    "Source: data.gov.sg",
    position = c("left", "bottom")
  )
```

Compass, scale bar, grid, and source note help the map read like a proper map output.

### Map Legend and Layout

```{r, fig.width=8, fig.height=7}
tm_shape(mpsz_pop2020) +
  tm_polygons(
    fill = "DEPENDENCY",
    fill.scale = tm_scale_intervals(
      style = "quantile",
      n = 5
    ),
    fill.legend = tm_legend(
      title = "Dependency ratio"
    )
  ) +
  tm_pos_auto_in() +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type = "8star", size = 2) +
  tm_scalebar() +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits(
    "Source: data.gov.sg",
    position = c("left", "bottom")
  )
```

The legend title makes the mapped variable clearer. This is a small change, but it helps the reader.

### Map Style

```{r, fig.width=8, fig.height=7}
tm_shape(mpsz_pop2020) +
  tm_polygons(
    fill = "DEPENDENCY",
    fill.scale = tm_scale_intervals(
      style = "quantile",
      n = 5,
      values = "-brewer.greens"
    )
  ) +
  tm_borders(fill_alpha = 0.5) +
  tmap_style("natural")
```

```{r}
tmap_style("white")
```

The style changes the overall look of the map. I reset it back to white after this demo.

### Small Multiple Maps by Multiple Fill Variables

```{r, fig.width=10, fig.height=6}
tm_shape(mpsz_pop2020) +
  tm_polygons(
    fill = c("YOUNG", "AGED"),
    fill.legend = tm_legend(
      position = tm_pos_in("right", "bottom")
    ),
    fill.scale = tm_scale_intervals(
      style = "equal",
      n = 5,
      values = "brewer.blues"
    )
  ) +
  tm_borders(fill_alpha = 0.5) +
  tmap_style("natural")
```

This creates small multiples from two variables. It is useful when I want to compare young and aged population patterns.

```{r}
tmap_style("white")
```

### Small Multiple Maps with tmap_arrange

```{r, fig.width=10, fig.height=6}
youngmap <- tm_shape(mpsz_pop2020) +
  tm_polygons(
    fill = "YOUNG",
    fill.legend = tm_legend(
      position = tm_pos_in("right", "bottom"),
      item.height = 0.8
    ),
    fill.scale = tm_scale_intervals(
      style = "quantile",
      values = "brewer.blues"
    )
  ) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Distribution of young population")

agedmap <- tm_shape(mpsz_pop2020) +
  tm_polygons(
    fill = "AGED",
    fill.legend = tm_legend(
      position = tm_pos_in("right", "bottom"),
      item.height = 0.8
    ),
    fill.scale = tm_scale_intervals(
      style = "quantile",
      values = "brewer.blues"
    )
  ) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Distribution of aged population")

tmap_arrange(youngmap, agedmap, asp = 1, ncol = 2)
```

This version builds two separate maps first. Then I arrange them side by side.

### Small Multiple Maps with tm_facets

```{r, fig.width=10, fig.height=7}
tm_shape(mpsz_pop2020) +
  tm_fill(
    fill = "DEPENDENCY",
    fill.scale = tm_scale_intervals(
      style = "quantile",
      values = "brewer.blues"
    )
  ) +
  tm_facets(
    by = "REGION_N",
    nrow = 2,
    ncols = 3,
    free.coords = TRUE,
    drop.units = TRUE
  ) +
  tm_layout(
    legend.show = TRUE,
    title.position = c("center", "center"),
    title.size = 20
  ) +
  tm_borders(fill_alpha = 0.5)
```

Facets split the map by region. This makes regional comparison easier.

### Select One Region

```{r, fig.width=8, fig.height=6}
mpsz_pop2020 %>%
  filter(REGION_N == "CENTRAL REGION") %>%
  tm_shape() +
  tm_polygons(
    fill = "DEPENDENCY",
    fill.scale = tm_scale_intervals(
      style = "quantile",
      values = "brewer.greens"
    ),
    fill.legend = tm_legend()
  ) +
  tm_borders(fill_alpha = 0.5)
```

This map focuses only on the Central Region. It is clearer when the study area is small.

### Add Statistical Chart and Outlier Highlight

```{r, fig.width=8, fig.height=6}
mpsz_selected <- mpsz_pop2020 %>%
  filter(REGION_N == "CENTRAL REGION")

mpsz_selected %>%
  tm_shape() +
  tm_polygons(
    fill = "DEPENDENCY",
    fill.scale = tm_scale_intervals(
      style = "quantile",
      values = "brewer.greens"
    ),
    fill.legend = tm_legend(),
    fill.chart = tm_chart_box()
  ) +
  tm_borders() +
  tm_layout(asp = 0.8)
```

The box chart gives a quick statistical view beside the map. It helps me see whether the mapped values have unusual cases.

```{r}
stats <- boxplot.stats(mpsz_selected$DEPENDENCY)
outlier_vals <- stats$out

outlier_vals
```

```{r}
outlier_sf <- mpsz_selected[mpsz_selected$DEPENDENCY %in% outlier_vals, ]

outlier_sf %>%
  st_drop_geometry() %>%
  select(SUBZONE_N, PLN_AREA_N, REGION_N, DEPENDENCY) %>%
  kable()
```

```{r, fig.width=8, fig.height=6}
tm_shape(mpsz_selected) +
  tm_polygons(
    fill = "DEPENDENCY",
    fill.scale = tm_scale_intervals(
      style = "quantile",
      values = "brewer.blues"
    ),
    fill.legend = tm_legend(),
    fill.chart = tm_chart_box()
  ) +
  tm_borders(fill_alpha = 0.5) +
  tm_shape(outlier_sf) +
  tm_borders(col = "red", lwd = 2) +
  tm_text("SUBZONE_N", col = "red", size = 0.7) +
  tm_layout(asp = 0.8)
```

The red border highlights outliers. This makes the statistical result visible on the map.

### Interactive Map

```{r}
region_selected <- mpsz_pop2020 %>%
  filter(REGION_N == "CENTRAL REGION")

region_bbox <- st_bbox(region_selected)

stats <- boxplot.stats(region_selected$DEPENDENCY)
outlier_vals <- stats$out
outlier_sf <- region_selected[region_selected$DEPENDENCY %in% outlier_vals, ]
```

```{r, fig.width=9, fig.height=6}
tmap_mode("view")

tm_shape(
  region_selected,
  bbox = region_bbox
) +
  tm_fill(
    "DEPENDENCY",
    id = "SUBZONE_N",
    popup.vars = c(
      "Name" = "SUBZONE_N",
      "Dependency" = "DEPENDENCY"
    )
  ) +
  tm_borders() +
  tm_shape(outlier_sf) +
  tm_borders(col = "red", lwd = 2)
```

This map is interactive. I can zoom, pan, and click a subzone to see the popup information.

```{r, fig.width=9, fig.height=6}
tm_shape(
  region_selected,
  bbox = region_bbox
) +
  tm_fill(
    "DEPENDENCY",
    id = "SUBZONE_N",
    popup.vars = c(
      "Name" = "SUBZONE_N",
      "Dependency" = "DEPENDENCY"
    )
  ) +
  tm_borders() +
  tm_shape(outlier_sf) +
  tm_borders(col = "red", lwd = 2) +
  tm_view(set_zoom_limits = c(12, 14))
```

The zoom limit keeps the reader near the study area. This makes the interactive map easier to control.

```{r}
tmap_mode("plot")
```

## Final Reflection

This hands-on exercise shows the full flow from raw data to maps. First, I prepare the spatial data and check the CRS. Then I create useful variables such as buffer area, pre-school density, and dependency ratio. Finally, I use different tmap options to show how map design can change the way a pattern is read.
