---
title: "Hands-on Exercise 3: Spatio-Temporal Point Patterns Analysis"
execute:
  echo: true
  warning: false
  message: false
format:
  html:
    toc: true
    toc-depth: 3
---

## Overview

This page follows [Chapter 6: Spatio-Temporal Point Patterns Analysis](https://r4gdsa.netlify.app/chap06).

The chapter studies forest fire points in Kepulauan Bangka Belitung, Indonesia. To keep the data size small, I use a light subset of Indonesia 2023 active fire detections and only keep the records around Bangka Belitung.

The main question is simple:

-   Are fire points randomly distributed in space and time?
-   If not, where and when do they cluster?

::: callout-note
The full chapter uses `sparr::spattemp.density()` and `stpp`. For this submission, I use a lighter `spatstat` workflow so the page can render reliably. The idea is the same: combine point locations with their event time.
:::

## Getting Started

### Load Packages

```{r}
library(sf)
library(spatstat)
library(tmap)
library(tidyverse)
library(lubridate)
library(knitr)
library(rnaturalearth)

set.seed(1234)
tmap_mode("plot")
```

```{r}
tibble(
  package = c("sf", "spatstat", "tmap", "tidyverse", "lubridate", "rnaturalearth"),
  version = c(
    as.character(packageVersion("sf")),
    as.character(packageVersion("spatstat")),
    as.character(packageVersion("tmap")),
    as.character(packageVersion("tidyverse")),
    as.character(packageVersion("lubridate")),
    as.character(packageVersion("rnaturalearth"))
  )
) %>%
  kable()
```

`sf` is used for spatial data. `spatstat` is used for point pattern objects and KDE. `tmap` is used for map display.

### Data Used

```{r}
data_used <- tibble(
  data_type = c("Geospatial", "Aspatial"),
  file = c(
    "Indonesia country boundary, clipped to Bangka Belitung area",
    "Indonesia 2023 active fire detections, filtered to Bangka Belitung area"
  ),
  path = c(
    "rnaturalearth::ne_countries(country = 'Indonesia')",
    "data/aspatial/indonesia_fires_2023_bangka_belitung.csv"
  ),
  source = c(
    "Natural Earth through rnaturalearth",
    "NASA FIRMS data, light CSV subset"
  )
)

kable(data_used)
```

I do not use the full Indonesia fire data in the report. The saved CSV only contains points near the study area.

## Part 1: Importing and Preparing Study Area

The study area is Bangka Belitung in Indonesia. I use the Indonesia country outline from Natural Earth and crop it to the Bangka Belitung bounding box.

### Create Study Area Boundary

```{r}
kbb_bbox <- st_as_sfc(
  st_bbox(
    c(
      xmin = 105.05,
      ymin = -3.20,
      xmax = 106.95,
      ymax = -1.40
    ),
    crs = 4326
  )
)

indonesia <- ne_countries(
  country = "Indonesia",
  scale = "small",
  returnclass = "sf"
)

indonesia_utm <- st_transform(st_make_valid(indonesia), 32748)
kbb_bbox_utm <- st_transform(kbb_bbox, 32748)

kbb_area <- suppressWarnings(
  st_intersection(indonesia_utm, kbb_bbox_utm)
) %>%
  st_geometry() %>%
  st_union() %>%
  st_make_valid()

kbb_area_sf <- st_as_sf(st_sfc(kbb_area, crs = 32748))
```

```{r}
tibble(
  geometry_type = paste(unique(st_geometry_type(kbb_area_sf)), collapse = ", "),
  epsg = st_crs(kbb_area_sf)$epsg,
  area_sqkm = as.numeric(st_area(kbb_area_sf)) / 1000000
) %>%
  mutate(area_sqkm = round(area_sqkm, 2)) %>%
  kable()
```

EPSG:32748 is UTM Zone 48S. It is suitable here because Bangka Belitung is in southern Indonesia and the units are metres.

### Convert Study Area to owin

```{r}
kbb_owin <- as.owin(kbb_area)

kbb_owin
```

```{r}
class(kbb_owin)
```

The `owin` object is the observation window. It tells `spatstat` which part of space belongs to the study area.

```{r, fig.width=7, fig.height=5}
plot(kbb_owin, main = "Bangka Belitung study window")
```

## Part 2: Importing and Preparing Fire Points

The fire data has longitude, latitude, date, time, satellite, confidence and fire radiative power. I convert it into an `sf` point layer first.

### Import Fire Data

```{r}
fire_raw <- read_csv(
  "data/aspatial/indonesia_fires_2023_bangka_belitung.csv",
  show_col_types = FALSE
)

glimpse(fire_raw)
```

```{r}
fire_sf_raw <- fire_raw %>%
  mutate(
    acq_date = as.Date(acq_date),
    DayofYear = yday(acq_date),
    Month_num = month(acq_date),
    Month_fac = factor(month.name[Month_num], levels = month.name)
  ) %>%
  st_as_sf(
    coords = c("longitude", "latitude"),
    crs = 4326,
    remove = FALSE
  ) %>%
  st_transform(crs = 32748)

inside_kbb <- lengths(st_intersects(fire_sf_raw, kbb_area_sf)) > 0
fire_sf <- fire_sf_raw[inside_kbb, ]
```

```{r}
fire_summary <- tibble(
  records_in_csv = nrow(fire_raw),
  records_in_study_area = nrow(fire_sf),
  first_date = min(fire_sf$acq_date),
  last_date = max(fire_sf$acq_date),
  mean_frp = mean(fire_sf$frp, na.rm = TRUE),
  max_frp = max(fire_sf$frp, na.rm = TRUE)
) %>%
  mutate(
    mean_frp = round(mean_frp, 2),
    max_frp = round(max_frp, 2)
  )

kable(fire_summary)
```

The CSV is already a subset. I do one more spatial filter so only points inside the approximate Bangka Belitung land boundary are analysed.

### Overall Fire Point Map

```{r, fig.width=8, fig.height=6}
tm_shape(kbb_area_sf) +
  tm_polygons(fill = "grey95", col = "grey55") +
  tm_shape(fire_sf) +
  tm_dots(size = 0.04, col = "#d95f0e", alpha = 0.55) +
  tm_title("Active Fire Detections in Bangka Belitung, 2023")
```

The fire detections are not evenly spread. Most points are concentrated in several parts of the islands.

### Monthly Fire Counts

```{r}
month_counts <- fire_sf %>%
  st_drop_geometry() %>%
  count(Month_fac, name = "fire_points") %>%
  complete(
    Month_fac = factor(month.name, levels = month.name),
    fill = list(fire_points = 0)
  )

kable(month_counts)
```

```{r, fig.width=8, fig.height=4}
ggplot(month_counts, aes(x = Month_fac, y = fire_points)) +
  geom_col(fill = "#d95f0e") +
  labs(
    title = "Monthly Active Fire Detections",
    x = NULL,
    y = "Number of points"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The peak is in September and October. This tells me that time is important, not only location.

### Fire Points by Month

```{r, fig.width=10, fig.height=8}
tm_shape(kbb_area_sf) +
  tm_polygons(fill = "grey95", col = "grey70") +
  tm_shape(fire_sf) +
  tm_dots(size = 0.05, col = "#de2d26", alpha = 0.55) +
  tm_facets(
    by = "Month_fac",
    nrow = 3,
    ncol = 4,
    free.coords = FALSE,
    drop.units = TRUE
  )
```

The monthly maps show a clear seasonal pattern. Early months have few detections, while September and October have many.

## Part 3: Creating Spatio-Temporal Point Objects

For `spatstat`, I need a `ppp` object. The location is stored as x and y coordinates, and the month or day is stored as a mark.

### Create ppp Object with Month Marks

```{r}
make_fire_ppp <- function(mark_values) {
  xy <- st_coordinates(fire_sf)

  fire_ppp <- ppp(
    x = xy[, "X"],
    y = xy[, "Y"],
    window = as.owin(st_bbox(kbb_area_sf)),
    marks = mark_values
  )

  fire_ppp[kbb_owin]
}

fire_month_owin <- make_fire_ppp(fire_sf$Month_num)
marks(fire_month_owin) <- factor(
  marks(fire_month_owin),
  levels = 1:12,
  labels = month.name
)

fire_month_owin
```

```{r}
fire_ppp_summary <- tibble(
  points = npoints(fire_month_owin),
  window_type = Window(fire_month_owin)$type,
  window_area_sqkm = spatstat.geom::area.owin(Window(fire_month_owin)) / 1000000,
  duplicated_points = sum(duplicated(fire_month_owin))
) %>%
  mutate(window_area_sqkm = round(window_area_sqkm, 2))

kable(fire_ppp_summary)
```

This object keeps both parts of the problem. The x-y location is spatial, and the mark is temporal.

### Create ppp Object with Day-of-Year Marks

```{r}
fire_yday_owin <- make_fire_ppp(fire_sf$DayofYear)

fire_yday_owin
```

```{r}
tibble(
  min_day = min(marks(fire_yday_owin)),
  median_day = median(marks(fire_yday_owin)),
  mean_day = mean(marks(fire_yday_owin)),
  max_day = max(marks(fire_yday_owin))
) %>%
  mutate(across(where(is.numeric), ~ round(.x, 1))) %>%
  kable()
```

Day-of-year keeps more time detail than month. It is useful for looking at shorter fire seasons.

## Part 4: STKDE by Month

The course computes spatio-temporal KDE. Here I use a lightweight version: split the points by month and compute a KDE for each month using the same spatial bandwidth.

### Split Points by Month

```{r}
fire_by_month <- split(fire_month_owin)
focus_months <- c("July", "August", "September", "October", "November", "December")
space_bandwidth <- 15000
```

```{r}
month_kde_stats <- map_dfr(focus_months, function(current_month) {
  month_ppp <- unmark(fire_by_month[[current_month]])
  month_kde <- density(
    month_ppp,
    sigma = space_bandwidth,
    edge = TRUE
  )

  tibble(
    month = current_month,
    points = npoints(month_ppp),
    max_density = max(month_kde$v, na.rm = TRUE)
  )
}) %>%
  mutate(max_density = signif(max_density, 3))

kable(month_kde_stats)
```

I use the same 15 km bandwidth for all months. This makes the monthly density maps easier to compare.

### Plot Monthly KDE Surfaces

```{r, fig.width=10, fig.height=7}
month_kde_list <- map(focus_months, function(current_month) {
  month_ppp <- unmark(fire_by_month[[current_month]])
  density(month_ppp, sigma = space_bandwidth, edge = TRUE)
})
names(month_kde_list) <- focus_months

old_par <- par(mfcol = c(2, 3), mar = c(2, 2, 2, 1))
for (current_month in focus_months) {
  plot(
    month_kde_list[[current_month]],
    main = paste("KDE at", current_month)
  )
}
par(old_par)
```

The strongest monthly surface appears around September and October. This matches the count table.

## Part 5: STKDE by Day of Year

Month is useful, but it is also coarse. I use day-of-year and a temporal weighting idea to create a few smoother time slices.

### Temporal Weighted KDE Function

```{r}
st_kde_slice <- function(target_day, lambda_days = 14, sigma = space_bandwidth) {
  temporal_weights <- dnorm(
    (marks(fire_yday_owin) - target_day) / lambda_days
  )

  density(
    unmark(fire_yday_owin),
    sigma = sigma,
    weights = temporal_weights,
    edge = TRUE
  )
}
```

The spatial bandwidth is 15 km. The temporal bandwidth is 14 days. So each output surface is mainly influenced by fires near the selected date.

### Plot Day-of-Year KDE Slices

```{r}
target_dates <- as.Date(c(
  "2023-09-01",
  "2023-10-01",
  "2023-11-01",
  "2023-12-01"
))

target_days <- yday(target_dates)

day_kde_list <- map(target_days, st_kde_slice)
names(day_kde_list) <- format(target_dates, "%d %b")
```

```{r, fig.width=9, fig.height=7}
old_par <- par(mfrow = c(2, 2), mar = c(2, 2, 2, 1))
for (current_date in names(day_kde_list)) {
  plot(
    day_kde_list[[current_date]],
    main = paste("STKDE slice:", current_date)
  )
}
par(old_par)
```

The slices show how the hotspot changes through the peak season. October is the strongest period in this subset.

```{r}
slice_stats <- tibble(
  date = names(day_kde_list),
  day_of_year = target_days,
  max_density = map_dbl(day_kde_list, ~ max(.x$v, na.rm = TRUE))
) %>%
  mutate(max_density = signif(max_density, 3))

kable(slice_stats)
```

This table gives a simple numeric check for the plotted surfaces.

## Part 6: Simple Space-Time Interaction Check

The full `stpp` workflow computes a formal space-time K-function. Here I make a simpler check by counting point pairs that are close in both space and time.

### Pair Distance Summary

```{r}
xy <- st_coordinates(fire_sf)
day_vec <- fire_sf$DayofYear

spatial_dist <- as.matrix(dist(xy)) / 1000
temporal_dist <- abs(outer(day_vec, day_vec, "-"))

upper_index <- upper.tri(spatial_dist)
space_pairs_km <- spatial_dist[upper_index]
time_pairs_days <- temporal_dist[upper_index]

total_pairs <- length(space_pairs_km)

interaction_table <- expand_grid(
  space_km = c(5, 10, 20),
  time_days = c(7, 14, 30)
) %>%
  rowwise() %>%
  mutate(
    pair_count = sum(space_pairs_km <= space_km & time_pairs_days <= time_days),
    pair_share = pair_count / total_pairs
  ) %>%
  ungroup() %>%
  mutate(pair_share = round(pair_share, 4))

kable(interaction_table)
```

If many pairs are close in both space and time, it suggests a clustered space-time process. The share increases when the space and time windows become wider.

### Space-Time Distance Plot

```{r, fig.width=8, fig.height=5}
pair_sample <- tibble(
  space_km = space_pairs_km,
  time_days = time_pairs_days
)

pair_sample_size <- min(6000, nrow(pair_sample))
pair_sample <- slice_sample(pair_sample, n = pair_sample_size)

ggplot(pair_sample, aes(x = space_km, y = time_days)) +
  geom_bin2d(bins = 35) +
  scale_fill_viridis_c() +
  coord_cartesian(xlim = c(0, 120), ylim = c(0, 120)) +
  labs(
    title = "Sample of Space-Time Pair Distances",
    x = "Spatial distance between two fire points (km)",
    y = "Temporal distance between two fire points (days)",
    fill = "Pairs"
  ) +
  theme_minimal()
```

The lower-left part of the plot represents fire pairs that are close in both space and time. This supports the idea of space-time clustering during the fire season.

## Final Reflection

This hands-on exercise extends point pattern analysis from space into space-time. I first prepare the study window and fire points, then use month and day-of-year as temporal marks. The monthly KDE shows when the fire activity becomes stronger. The day-of-year weighted KDE gives smoother time slices. The pair-distance table is a simple check that many fires happen near other fires in both space and time. Overall, the pattern is not just spatial. It also changes clearly through the year.
