---
title: "Hands-on Exercise 2: Spatial Point Pattern Analysis"
execute:
  echo: true
  warning: false
  message: false
format:
  html:
    toc: true
    toc-depth: 3
---

## Overview

This page combines two Hands-on Exercise 2 topics:

-   [First-order Spatial Point Patterns Analysis Methods](https://r4gdsa.netlify.app/chap04.html)
-   [2nd Order Spatial Point Patterns Analysis Methods](https://r4gdsa.netlify.app/chap05.html)

The two chapters use the same child care point data. Chapter 4 focuses on where the points are dense. Chapter 5 checks whether the points are spatially random or show interaction at different distances.

## Getting Started

### Load Packages

```{r}
library(sf)
library(terra)
library(spatstat)
library(tmap)
library(tidyverse)
library(knitr)

set.seed(1234)
tmap_mode("plot")
```

```{r}
tibble(
  package = c("sf", "terra", "spatstat", "tmap", "tidyverse"),
  version = c(
    as.character(packageVersion("sf")),
    as.character(packageVersion("terra")),
    as.character(packageVersion("spatstat")),
    as.character(packageVersion("tmap")),
    as.character(packageVersion("tidyverse"))
  )
) %>%
  kable()
```

These packages cover the full workflow. `sf` handles vector data, `spatstat` handles point pattern analysis, and `tmap` is used for map output.

### Data Preparation

```{r}
data_used <- tibble(
  data_type = c("Geospatial", "Geospatial"),
  file = c(
    "Master Plan 2019 Subzone Boundary (No Sea)",
    "Child Care Services"
  ),
  path = c(
    "data/geospatial/master_plan_2019_subzone_no_sea.geojson",
    "data/geospatial/childcare_services.geojson"
  ),
  source = c("data.gov.sg", "data.gov.sg")
)

kable(data_used)
```

The boundary layer gives the study area. The child care layer gives the point events for the analysis.

## Part 1: First-order Spatial Point Patterns Analysis

First-order analysis is mainly about intensity. In simple terms, I am asking where child care services are more concentrated.

### Import and Transform Geospatial Data

```{r}
mpsz_sf <- st_read(
  "data/geospatial/master_plan_2019_subzone_no_sea.geojson",
  quiet = TRUE
) %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_transform(crs = 3414)

childcare_sf <- st_read(
  "data/geospatial/childcare_services.geojson",
  quiet = TRUE
) %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_transform(crs = 3414)
```

```{r}
layer_summary <- tibble(
  layer = c("mpsz_sf", "childcare_sf"),
  features = c(nrow(mpsz_sf), nrow(childcare_sf)),
  geometry_type = c(
    paste(unique(st_geometry_type(mpsz_sf)), collapse = ", "),
    paste(unique(st_geometry_type(childcare_sf)), collapse = ", ")
  ),
  epsg = c(st_crs(mpsz_sf)$epsg, st_crs(childcare_sf)$epsg)
)

kable(layer_summary)
```

EPSG:3414 uses metres. This matters because nearest neighbour distance and KDE bandwidth both depend on distance units.

### Map the Child Care Services

```{r, fig.width=8, fig.height=7}
tm_shape(mpsz_sf) +
  tm_borders(col = "grey70", lwd = 0.3) +
  tm_shape(childcare_sf) +
  tm_dots(size = 0.03, col = "#2b8cbe", alpha = 0.55) +
  tm_title("Child Care Services in Singapore")
```

The points are not evenly spread across the island. They are mostly in residential planning areas, which is expected for this type of facility.

### Convert sf Data to ppp

```{r}
coords <- st_coordinates(childcare_sf)

childcare_ppp <- ppp(
  x = coords[, "X"],
  y = coords[, "Y"],
  window = as.owin(st_bbox(mpsz_sf))
)

class(childcare_ppp)
```

The course uses `as.ppp(childcare_sf)`. In my current `spatstat` setup, I create the same `ppp` object explicitly from the point coordinates.

```{r}
tibble(
  total_points = npoints(childcare_ppp),
  duplicated_xy = sum(duplicated(as.data.frame(coords))),
  window_type = childcare_ppp$window$type
) %>%
  kable()
```

Some centres share the same coordinate. This can happen when several services are in the same building.

### Create the Singapore Study Window

```{r}
mpsz_cl <- st_union(st_geometry(mpsz_sf))
sg_owin <- as.owin(mpsz_cl)

class(sg_owin)
```

```{r, fig.width=7, fig.height=6}
plot(sg_owin, main = "Singapore study window")
```

The `owin` object tells `spatstat` where the study area is. Without it, the point pattern would only know the rectangular bounding box.

### Combine Points and Window

```{r}
childcareSG_ppp <- unmark(childcare_ppp[sg_owin])

childcareSG_ppp
```

```{r}
window_summary <- tibble(
  points = npoints(childcareSG_ppp),
  window_type = Window(childcareSG_ppp)$type,
  window_area_sqkm = spatstat.geom::area.owin(Window(childcareSG_ppp)) / 1000000,
  average_intensity_per_sqkm = intensity(childcareSG_ppp) * 1000000
) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))

kable(window_summary)
```

Now the point pattern is clipped to the Singapore boundary. This is the object used in the nearest neighbour and density analysis.

### Clark-Evans Test

```{r}
ce_global_z <- clarkevans.test(
  childcareSG_ppp,
  correction = "none",
  clipregion = NULL,
  alternative = "clustered"
)

ce_global_z
```

The Clark-Evans R value is below 1. This means the child care services are closer together than a random pattern.

```{r}
ce_global_mc <- clarkevans.test(
  childcareSG_ppp,
  correction = "none",
  clipregion = NULL,
  alternative = "clustered",
  method = "MonteCarlo",
  nsim = 99
)

ce_global_mc
```

The Monte Carlo test gives the same broad conclusion. The observed pattern is more clustered than Complete Spatial Randomness.

### Kernel Density Estimation

```{r, fig.width=7, fig.height=6}
kde_SG_diggle <- density(
  childcareSG_ppp,
  sigma = bw.diggle,
  edge = TRUE,
  kernel = "gaussian"
)

plot(kde_SG_diggle, main = "KDE of child care services, bw.diggle")
```

KDE turns point locations into a smooth surface. High values show places with stronger concentration of child care services.

```{r}
bw <- bw.diggle(childcareSG_ppp)
bw
```

The bandwidth controls how smooth the density surface is. A larger bandwidth makes a smoother map.

### Rescale KDE Values

```{r}
childcareSG_ppp_km <- rescale.ppp(
  childcareSG_ppp,
  1000,
  "km"
)
```

```{r, fig.width=7, fig.height=6}
kde_childcareSG_km <- density(
  childcareSG_ppp_km,
  sigma = bw.diggle,
  edge = TRUE,
  kernel = "gaussian"
)

plot(kde_childcareSG_km, main = "KDE after rescaling to kilometres")
```

Rescaling makes the distance unit easier to read. The pattern is the same, but the bandwidth is now expressed in kilometres.

### Compare Bandwidth Methods

```{r}
bw_scott <- bw.scott(childcareSG_ppp_km)

bandwidth_table <- tibble(
  method = c("bw.CvL", "bw.scott x", "bw.scott y", "bw.ppl", "bw.diggle"),
  sigma_km = c(
    as.numeric(bw.CvL(childcareSG_ppp_km)),
    as.numeric(bw_scott["sigma.x"]),
    as.numeric(bw_scott["sigma.y"]),
    as.numeric(bw.ppl(childcareSG_ppp_km)),
    as.numeric(bw.diggle(childcareSG_ppp_km))
  )
) %>%
  mutate(sigma_km = round(sigma_km, 3))

kable(bandwidth_table)
```

Different methods choose different bandwidths. I keep `bw.diggle` for most maps because it gives a clear local pattern without being too rough.

```{r, fig.width=10, fig.height=5}
kde_childcareSG_ppl <- density(
  childcareSG_ppp_km,
  sigma = bw.ppl,
  edge = TRUE,
  kernel = "gaussian"
)

old_par <- par(mfrow = c(1, 2))
plot(kde_childcareSG_km, main = "bw.diggle")
plot(kde_childcareSG_ppl, main = "bw.ppl")
par(old_par)
```

`bw.ppl` uses a larger bandwidth here, so the density surface is smoother. `bw.diggle` keeps more local detail.

### Compare Kernel Methods

```{r, fig.width=10, fig.height=8}
sigma_km <- as.numeric(bw.diggle(childcareSG_ppp_km))

old_par <- par(mfrow = c(2, 2))
plot(
  density(childcareSG_ppp_km, sigma = sigma_km, edge = TRUE, kernel = "gaussian"),
  main = "Gaussian"
)
plot(
  density(childcareSG_ppp_km, sigma = sigma_km, edge = TRUE, kernel = "epanechnikov"),
  main = "Epanechnikov"
)
plot(
  density(childcareSG_ppp_km, sigma = sigma_km, edge = TRUE, kernel = "quartic"),
  main = "Quartic"
)
plot(
  density(childcareSG_ppp_km, sigma = sigma_km, edge = TRUE, kernel = "disc"),
  main = "Disc"
)
par(old_par)
```

Changing the kernel changes the surface shape slightly. The main hotspot pattern remains similar.

### Fixed and Adaptive KDE

```{r, fig.width=10, fig.height=5}
kde_childcareSG_fb <- density(
  childcareSG_ppp_km,
  sigma = 0.6,
  edge = TRUE,
  kernel = "gaussian"
)

kde_childcareSG_ab <- adaptive.density(
  childcareSG_ppp_km,
  method = "kernel"
)

old_par <- par(mfrow = c(1, 2))
plot(kde_childcareSG_fb, main = "Fixed bandwidth, 0.6 km")
plot(kde_childcareSG_ab, main = "Adaptive bandwidth")
par(old_par)
```

Fixed bandwidth uses one smoothing distance everywhere. Adaptive bandwidth changes the smoothing by local point density.

### Cartographic KDE Map

```{r}
kde_childcareSG_terra <- terra::rast(kde_SG_diggle)
crs(kde_childcareSG_terra) <- "EPSG:3414"

kde_childcareSG_terra
```

I use the metre-based KDE here so the raster coordinates match EPSG:3414.

```{r, fig.width=8, fig.height=7}
tm_shape(kde_childcareSG_terra) +
  tm_raster(
    col.scale = tm_scale_continuous(values = "viridis"),
    col.legend = tm_legend(
      title = "Density values",
      title.size = 0.7,
      text.size = 0.7,
      bg.color = "white",
      bg.alpha = 0.75,
      position = tm_pos_in("right", "bottom"),
      frame = TRUE
    )
  ) +
  tm_compass() +
  tm_layout(scale = 1.0)
```

This is a cleaner map output. It shows the KDE result as a raster surface instead of a basic R plot.

### First-order Analysis by Planning Area

```{r}
make_area_window <- function(planning_area) {
  mpsz_sf %>%
    filter(PLN_AREA_N == planning_area) %>%
    st_geometry() %>%
    st_union() %>%
    as.owin()
}

make_area_area <- function(planning_area) {
  mpsz_sf %>%
    filter(PLN_AREA_N == planning_area) %>%
    st_geometry() %>%
    st_union() %>%
    st_area() %>%
    as.numeric()
}

pg_owin <- make_area_window("PUNGGOL")
tm_owin <- make_area_window("TAMPINES")
ck_owin <- make_area_window("CHOA CHU KANG")
jw_owin <- make_area_window("JURONG WEST")

childcare_pg_ppp <- unmark(childcare_ppp[pg_owin])
childcare_tm_ppp <- unmark(childcare_ppp[tm_owin])
childcare_ck_ppp <- unmark(childcare_ppp[ck_owin])
childcare_jw_ppp <- unmark(childcare_ppp[jw_owin])
```

```{r}
planning_area_summary <- tibble(
  planning_area = c("Punggol", "Tampines", "Choa Chu Kang", "Jurong West"),
  childcare_points = c(
    npoints(childcare_pg_ppp),
    npoints(childcare_tm_ppp),
    npoints(childcare_ck_ppp),
    npoints(childcare_jw_ppp)
  ),
  area_sqkm = c(
    make_area_area("PUNGGOL"),
    make_area_area("TAMPINES"),
    make_area_area("CHOA CHU KANG"),
    make_area_area("JURONG WEST")
  ) / 1000000
) %>%
  mutate(
    area_sqkm = round(area_sqkm, 2),
    points_per_sqkm = round(childcare_points / area_sqkm, 2)
  )

kable(planning_area_summary)
```

These four planning areas have enough points for local comparison. The point count alone is not enough, so I also calculate points per square kilometre.

```{r}
childcare_pg_ppp_km <- rescale.ppp(childcare_pg_ppp, 1000, "km")
childcare_tm_ppp_km <- rescale.ppp(childcare_tm_ppp, 1000, "km")
childcare_ck_ppp_km <- rescale.ppp(childcare_ck_ppp, 1000, "km")
childcare_jw_ppp_km <- rescale.ppp(childcare_jw_ppp, 1000, "km")
```

```{r, fig.width=10, fig.height=8}
old_par <- par(mfrow = c(2, 2))
plot(childcare_pg_ppp_km, main = "Punggol")
plot(childcare_tm_ppp_km, main = "Tampines")
plot(childcare_ck_ppp_km, main = "Choa Chu Kang")
plot(childcare_jw_ppp_km, main = "Jurong West")
par(old_par)
```

This local view makes the point pattern easier to inspect. The whole-island view hides many small differences.

### Clark-Evans Test by Planning Area

```{r}
ce_ck <- clarkevans.test(
  childcare_ck_ppp,
  correction = "none",
  clipregion = NULL,
  alternative = "two.sided"
)

ce_tm <- clarkevans.test(
  childcare_tm_ppp,
  correction = "none",
  clipregion = NULL,
  alternative = "two.sided"
)

ce_ck
ce_tm
```

```{r}
area_ce_table <- tibble(
  planning_area = c("Choa Chu Kang", "Tampines"),
  R = c(unname(ce_ck$statistic), unname(ce_tm$statistic)),
  p_value = format.pval(
    c(ce_ck$p.value, ce_tm$p.value),
    digits = 3,
    eps = 0.001
  )
) %>%
  mutate(R = round(R, 3))

kable(area_ce_table)
```

Both R values are below 1. This suggests clustering at the planning area level too.

### KDE by Planning Area

```{r, fig.width=10, fig.height=8}
old_par <- par(mfrow = c(2, 2))
plot(
  density(childcare_pg_ppp_km, sigma = bw.diggle, edge = TRUE, kernel = "gaussian"),
  main = "Punggol"
)
plot(
  density(childcare_tm_ppp_km, sigma = bw.diggle, edge = TRUE, kernel = "gaussian"),
  main = "Tampines"
)
plot(
  density(childcare_ck_ppp_km, sigma = bw.diggle, edge = TRUE, kernel = "gaussian"),
  main = "Choa Chu Kang"
)
plot(
  density(childcare_jw_ppp_km, sigma = bw.diggle, edge = TRUE, kernel = "gaussian"),
  main = "Jurong West"
)
par(old_par)
```

The KDE maps show local high-density areas. This is more useful than only saying whether the whole pattern is clustered.

## Part 2: Second-order Spatial Point Patterns Analysis

Second-order analysis looks at distance relationships between points. Here I use Choa Chu Kang and Tampines because they have enough child care points and the course also uses them.

::: callout-note
The course uses 999 simulations for some CSR tests. I use `simulation_n <- 99` so the page can render faster. The code structure is the same, and a larger number can be used for a final statistical run.
:::

```{r}
simulation_n <- 99
gf_rmax <- 500
kl_rmax <- 1000
```

### G-function

The G-function looks at nearest neighbour distance from each observed point. If the observed curve rises faster than the CSR curve, nearby points are common.

```{r, fig.width=10, fig.height=5}
G_CK <- Gest(childcare_ck_ppp, correction = "border", rmax = gf_rmax)
G_tm <- Gest(childcare_tm_ppp, correction = "best", rmax = gf_rmax)

old_par <- par(mfrow = c(1, 2))
plot(G_CK, xlim = c(0, gf_rmax), main = "G-function: Choa Chu Kang")
plot(G_tm, xlim = c(0, gf_rmax), main = "G-function: Tampines")
par(old_par)
```

The two plots show how quickly each child care centre finds another nearby centre.

```{r, fig.width=10, fig.height=5}
G_CK_csr <- envelope(
  childcare_ck_ppp,
  Gest,
  nsim = simulation_n,
  rmax = gf_rmax,
  verbose = FALSE
)

G_tm_csr <- envelope(
  childcare_tm_ppp,
  Gest,
  correction = "all",
  nsim = simulation_n,
  rmax = gf_rmax,
  verbose = FALSE
)

old_par <- par(mfrow = c(1, 2))
plot(G_CK_csr, main = "CSR envelope: Choa Chu Kang")
plot(G_tm_csr, main = "CSR envelope: Tampines")
par(old_par)
```

The envelope is the random benchmark. When the observed line moves outside the envelope, the pattern is not behaving like CSR at that distance.

### F-function

The F-function uses random empty locations and measures the distance to the nearest child care point. It helps describe empty space.

```{r, fig.width=10, fig.height=5}
F_CK <- Fest(childcare_ck_ppp, rmax = gf_rmax)
F_tm <- Fest(childcare_tm_ppp, correction = "best", rmax = gf_rmax)

old_par <- par(mfrow = c(1, 2))
plot(F_CK, main = "F-function: Choa Chu Kang")
plot(F_tm, main = "F-function: Tampines")
par(old_par)
```

If the F curve rises quickly, most places are close to a child care service. This is about coverage from the empty-space side.

```{r, fig.width=10, fig.height=5}
F_CK_csr <- envelope(
  childcare_ck_ppp,
  Fest,
  nsim = simulation_n,
  rmax = gf_rmax,
  verbose = FALSE
)

F_tm_csr <- envelope(
  childcare_tm_ppp,
  Fest,
  correction = "all",
  nsim = simulation_n,
  rmax = gf_rmax,
  verbose = FALSE
)

old_par <- par(mfrow = c(1, 2))
plot(F_CK_csr, main = "CSR envelope: Choa Chu Kang")
plot(F_tm_csr, main = "CSR envelope: Tampines")
par(old_par)
```

This adds a random reference to the empty-space analysis. It is useful because a visual curve alone can be hard to judge.

### K-function

The K-function counts neighbours within increasing distance bands. It is useful for checking clustering across many distances.

```{r, fig.width=10, fig.height=5}
K_ck <- Kest(childcare_ck_ppp, correction = "Ripley", rmax = kl_rmax)
K_tm <- Kest(childcare_tm_ppp, correction = "Ripley", rmax = kl_rmax)

old_par <- par(mfrow = c(1, 2))
plot(K_ck, . - r ~ r, ylab = "K(d)-r", xlab = "d(m)", main = "K-function: Choa Chu Kang")
plot(K_tm, . - r ~ r, ylab = "K(d)-r", xlab = "d(m)", main = "K-function: Tampines")
par(old_par)
```

The transformed plot makes it easier to compare the observed pattern with the CSR reference line.

```{r, fig.width=10, fig.height=5}
K_ck_csr <- envelope(
  childcare_ck_ppp,
  Kest,
  nsim = simulation_n,
  rank = 1,
  global = TRUE,
  rmax = kl_rmax,
  verbose = FALSE
)

K_tm_csr <- envelope(
  childcare_tm_ppp,
  Kest,
  nsim = simulation_n,
  rank = 1,
  global = TRUE,
  rmax = kl_rmax,
  verbose = FALSE
)

old_par <- par(mfrow = c(1, 2))
plot(K_ck_csr, . - theo ~ r, main = "K CSR: Choa Chu Kang")
plot(K_tm_csr, . - theo ~ r, main = "K CSR: Tampines")
par(old_par)
```

For K-function envelopes, a line above the random envelope suggests clustering at that distance range.

### L-function

The L-function is a transformed version of K. It is often easier to read because CSR is shown around zero after subtracting distance.

```{r, fig.width=10, fig.height=5}
L_ck <- Lest(childcare_ck_ppp, correction = "Ripley", rmax = kl_rmax)
L_tm <- Lest(childcare_tm_ppp, correction = "Ripley", rmax = kl_rmax)

old_par <- par(mfrow = c(1, 2))
plot(L_ck, . - r ~ r, ylab = "L(d)-r", xlab = "d(m)", main = "L-function: Choa Chu Kang")
plot(L_tm, . - r ~ r, ylab = "L(d)-r", xlab = "d(m)", main = "L-function: Tampines")
par(old_par)
```

L-function is easier to explain visually. Values above zero usually indicate clustering.

```{r, fig.width=10, fig.height=5}
L_ck_csr <- envelope(
  childcare_ck_ppp,
  Lest,
  nsim = simulation_n,
  rank = 1,
  global = TRUE,
  rmax = kl_rmax,
  verbose = FALSE
)

L_tm_csr <- envelope(
  childcare_tm_ppp,
  Lest,
  nsim = simulation_n,
  rank = 1,
  global = TRUE,
  rmax = kl_rmax,
  verbose = FALSE
)

old_par <- par(mfrow = c(1, 2))
plot(L_ck_csr, . - theo ~ r, main = "L CSR: Choa Chu Kang")
plot(L_tm_csr, . - theo ~ r, main = "L CSR: Tampines")
par(old_par)
```

The CSR envelopes help confirm that the pattern is not only a visual impression. The services show clustering at short distances, especially in denser residential parts.

## Final Reflection

This hands-on exercise shows two ways to study point patterns. First-order methods show where the intensity is high. Second-order methods check how points relate to other points across distance. For child care services, clustering is expected because centres follow housing estates and population demand. The main limitation is that CSR is a simple benchmark, so the result should be read as an exploratory test rather than a full explanation of why centres are located there.
